Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-10-17 paper/introduction.tex
The type system ensures complete memory safety.

  \item \name is equipped with a region type system that statically
  guarantees safety for a well-defined subset of all memory accesses
  in a well-typed program. This subset includes all memory accesses,
  including accesses that happen across different regions, except
  dereferences of \C{Region} objects themselves. For the later kind,
  we rely on \GK{reference counting?} to ensure safety at run-time.
  Our experiments with \naiad workloads show that \GK{x\%} of all
  memory memory accesses fall under the statically safe subset, while
  only \GK{100-x\%} require run-time mechanisms to ensure safety.

>>>>

  \item \name is equipped with a region type system that statically
  guarantees safety all memory accesses in a well-typed program,
  subject to some conditions that can be checked efficiently at
  run-time. The overhead of checking these conditions is less than
  \GK{x\%} in our experiments with \naiad workloads.

### 2015-10-17 paper/overview.tex
The type system ensures complete memory safety.

A transferable region is said to be in \emph{closed} state when
created. \name provides \C{openalloc} lexical construct to \emph{open}
a transferable region as an allocation context, and to access its root
object:
\begin{codejava}
  Region<string> rgn = new Region<string>
                        (() => "Hello");
  openalloc rgn withroot str {
    print str; //prints "Hello"
    str = str + " World!"; 
  }
\end{codejava}
In the above code, the \C{openalloc} lexical block opens \C{rgn} for
allocation, and sets the newly introduced local variable (\C{str}) to
refer to its root object.  The assignment statement within the block
updates the root. Since \C{rgn} is the allocation context within the
block, the new string is allocated inside \C{rgn}, hence the
assignment is safe. In some cases, it is desirable to open a
transferable region only to read its contents, but not to perform any
allocations. For this purpose, \name provides \C{open} lexical
construct that, like \C{openalloc}, opens a transferable region and
names its root object, but unlike \C{openalloc}, does not make it an
allocation context. A transferable region can be \C{openalloc}'d or
\C{open}'d many times, before it is eventually \C{free}'d or
\C{transfer}'d.

The design decisions reflected above, \emph{viz.}, to ascribe a root
object to every transferable region, to make use of a higher-order
argument to initialize the root object, and to require that a region
be explictly opened in a lexical block for allocations, were made to
reflect the message semantics of transferable regions, to promote
uniform style of programming with different kinds of regions, and to
precisely state and enforce the safety guarantees offered by \name's
region type system (\S~\ref{sec:type-system}).

>>>>

\name imposes a simple discipline on how $\RgnZ$ objects can be used
to perform operations on transferable regions. This discipline is
imposed in the interest of memory safety, which is discussed in the
following subsection.

\subsection{Memory Safety}
\label{sec:memory-safety}

Memory safety in \name is primarily due to its region type system,
although few runtime checks are needed.
% However, the type system does not ensure the integrity of a small
% subset of pointers, called the \emph{unsafe set}, for which runtime
% verification is needed to guarantee safety. This is more a design
% decision rather than an unfortunate inevitability, as the following
% discussion elaborates.
At a fundamental level, the region type system's approach to memory
safety involves approximating (a). the set of memory regions that are
live at any given program location\footnote{We say the type system
tracks \emph{liveness} of memory regions}, and (b).  \emph{outlives}
relationships between them, and using this information to preempt
potentially unsafe references from being created. These include:
\begin{itemize}
\item References into dead/non-existing regions,
\item References from longer-living stack regions to shorter-living ones, and 
\item References that escape transferable regions.
\end{itemize}
Maintaining precise approximation of liveness and outlives information
for stack regions is easy. However, for dynamic (transferable) regions,
statically determining liveness is equivalent to determining if a C
pointer is null in presence of dynamic memory allocation, which is
undecidable in general. Systems that attempt to ensure full memory
safety in presence of dynamic memory (e.g., ~\cite{henglein},
~\cite{rust}) do so by conservatively approximating liveness
information for memory (regions). Maintaining a good conservative
approximation requires them to employ techniques such as linear
types~\cite{wadlerlt} or unique pointers~\cite{rust}, which impose
restrictions that are often too hard to program
around\footnote{Internet has many tutorials on how to \emph{fight} the
borrow checker in Rust.}. This is particularly unacceptable in the
context of \name, as dataflow programs are often run offline, and the
concern for performance and developer productivity outweighs the
concern for safety.

\name's region type system therefore adopts the novel approach of
overapproximating liveness information for transferable regions, i.e.,
it might assume that a transferable region is live when, in fact, it
might have already been transferred or freed. This imprecision should
be a cause of concern as it can lead to cases where creation of unsafe
references, and their subsquent dereferencing go unnoticed by the type
system. Considering that stack regions can contain pointers into
transferable regions, the worst case cost of overapproximating
liveness is therefore the integrity of \emph{every} pointer on stack.

Fortunately, \name avoids this worst case scenario.  \name achieves
this by imposing a simple state transition discipline on the lifetime
of a transferable region.  The relevant finite state machine is shown
in Fig.~\ref{fig:region-fsm}.

\begin{figure}
\includegraphics[scale=0.45]{region-fsm.png}
\caption{The lifetime of a dynamic (transferable) region in \name}
\label{fig:region-fsm}
\end{figure}

A transferable region starts its lifetime in a \emph{closed} state,
when it is created through \C{new Region} expresion. \name provides
\C{open} lexical construct to \emph{open} a transferable region as an
allocation context, and to access its root object:
\begin{codejava}
  Region<string> rgn = new Region<List<String>>
                        (() => new List<String>());
  open rgn as R0 withroot strList {
    strList.addAtHead("World");
    strList.addAtHead("Hello");
  }
  rgn.transfer();
\end{codejava}
In the above code, the \C{open} lexical block performs the
following operations: (a). It opens the transferable region handled by
\C{rgn} for allocation (i.e., makes it the current allocation
context), (b). assigns the identifier \C{R0} to the open region, and
(c). sets the newly introduced local variable (\C{strList}) to refer
to its root object (a list of strings). As Fig.~\ref{fig:region-fsm}
indicates, it is not possible to open a region that is already
transferred/freed. The fact that \C{rgn} is open within the block
therefore guarantees that it is not yet transferred/freed, and that
dereferencing \C{strList} within the block is safe. The end of
\C{open} block marks the return of transferable region to the closed
state.  While closed, the region is eligible to be transferred to a
downstream actor, or to be freed.  An actor that receives the
transferable region, receives it in the closed state. It can then
reopen the received region to read its contents, possibly add more
data and transfer it to another actor, or free the region.

\name lets stack regions to be used as working memory while operating
with the data stored in a transferable region. Consider the following
code, for instance\footnote{For brevity, we drop \C{as} part of the
\C{open} expression whenever there is no need to assign a static
identifier for the open region.}:
\begin{codejava}
void onReceive(Region<List<String>> rgn)
  open rgn withroot strList {
    letregion R1 {
      String s = "";
      ListIterator<String> i = strList.listIterator();
      while(i.hasNext()) {
        s += i.getNext();
      }
      print s; //prints "HelloWorld"
    }
  }
  rgn.free();
}
\end{codejava}
The stack region \C{R1} is being used in the above code to provide
working memory to work with the objects of transferable region
\C{rgn}. Since a transferable region cannot be transferred/freed while
it is still open (Fig~\ref{fig:region-fsm}), \C{rgn} is guaranteed to
outlive the stack region \C{R1} in the above code, making it safe for
the later to contain references to the former. \name therefore allows
such references.

% The design decisions reflected above, \emph{viz.}, to ascribe a root
% object to every transferable region, to make use of a higher-order
% argument to initialize the root object, and to require that a region
% be explictly opened in a lexical block for allocations, were made to
% reflect the message semantics of transferable regions, to promote
% uniform style of programming with different kinds of regions, and to
% precisely state and enforce the safety guarantees offered by \name's
% region type system (\S~\ref{sec:type-system}).

As evident in the above examples, the safety of memory accesses in
\name is now subject to the condition that the lifetime of every
transferable region aligns with the state transition discipline shown
in Fig.~\ref{fig:region-fsm}. If the transition discipline is somehow
guaranteed, then \name's region type system statically guarantees the
safety of all memory accesses. In other words, the type system reduces
the problem of ensuring memory safety in \name programs to the problem
of enforcing state transition discipline on the lifetimes of
transferable regions. In \name, this enforcement is done at runtime by
explicitly keeping track \emph{current state} for $\RgnZ$ objects, and
preventing them from going through invalid transitions.
